No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-09-26 paper/GEB Submission/manuscript_GEB.Rmd
Finish line edits

Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g. coral bleaching, @sully2019].
At the local scale however, species diversity tells a different story.
Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate [@boënnec2024; @johnson2024].
Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, `r colorize("short time windows biased toward the present day", "blue")` and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole, leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].

@@

We estimated general trends across bird and mammal communities for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, methods which deal well with the inherent imbalances in our data.
`r colorize("We fit individual models for each of the 18 CWM's calculated for continuous traits or trait categories (see diet, foraging strata, and body mass traits in Table 1).", "blue")` All time series with data for a given trait were included in the corresponding *CWM* model. We estimated study level trends using individual linear models. For studies with more than one location we fit a random slope and intercept for location. Study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models. They represented 13 of 1350 study-level models fit for each metric. For further details see the supplement. Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality. `r colorize("The log transformation was applied to all metrics except the SES corrected functional diversity metrics, FEve, and CWM's for`invertebrate`consumption, ground foraging strata, and below water surface foraging strata.", "blue")`

'r colorize("We tested for category-specific trends in species and functional diversity within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

We did find significant category-specific trends for Jaccard similarity and two dietary *CWM's*.
While Jaccard similarity was decreasing in the general trend and there were significant category-specific trends for birds, terrestrial, and temperate studies, there was no significant trend for marine, mammal, or tropical studies.
We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
We found significant dietary shifts across communities, with a significant general trend of increasing fish consumption, which was also reflected in increasing fish consumption trends for bird studies and temperate studies.
We found significantly decreasing trend for vertebrate consumption in marine studies and tropical studies.
The trend for marine vertebrate consumption was significantly more negative than for terrestrial studies and the temperate trend was significantly more negative than tropical studies.

At the study level, 11 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
For observed functional diversity metrics, 11 of 50 studies exhibited a trend in at least one metric, and 7 of 50 studies exhibited a significant trend for at least one corrected metric (Table \ref{tab:trendTab}).
In general, there were more significant trends for observed metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
Study-level slopes were not significantly related to start year or duration of time series for any metric.

@@

We assessed the sensitivity of general trend results to major data processing decision by rerunning models with increasingly conservative subsets of the data.
Trends for Jaccard similarity and fish consumption were not sensitive to either time series duration or trait coverage.
After excluding time series with less than three years we found an increasing trend for body mass that remained after excluding time series of less than four and five years.
The body size trend did however disappear after excluding time series of less then 10 years.
A complete list of models run in the sensitivity analysis and their results can be found in the supplement.

@@

Category-specific trends in `r colorize("species", "blue")` turnover highlight the types of communities experiencing change in community composition.
For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.
We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.

Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact.
For example, we found no trend for mean body mass in bird or mammal communities.
In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size.
This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts[@hantak2021], or current shifts are happening at an intraspecific level not captured by our data.
Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].

>>>>

Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g. coral bleaching, @sully2019].
At the local scale however, species diversity tells a different story.
Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate [@boënnec2024; @johnson2024].
Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, `r colorize("short time windows biased toward the present day", "blue")` and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole [@chapman2024], leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].

@@

We estimated general trends across bird and mammal communities for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, methods which deal well with the inherent imbalances in our data.
`r colorize("We fit individual models for each of the 18 CWM's calculated for continuous traits or trait categories (see diet, foraging strata, and body mass traits in Table 1).", "blue")` All time series with data for a given trait were included in the corresponding *CWM* model. We estimated study level trends using individual linear models. For studies with more than one location we fit a random slope and intercept for location. Study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models. They represented 13 of 1350 study-level models fit for each metric. For further details see the supplement. Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality. `r colorize("The log transformation was applied to all metrics except the SES corrected functional diversity metrics, FEve, and CWM's for invertebrate consumption, ground foraging strata, and below water surface foraging strata.", "blue")`

'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

@@

We did find significant category-specific trends for Jaccard similarity and two dietary *CWM's*.
While Jaccard similarity was decreasing in the general trend and there were significant category-specific `r colorize("turnover", "blue")` trends for birds, terrestrial, and temperate studies, there was no significant trend `r colorize("in Jaccard similarity", "blue")` for marine, mammal, or tropical studies.
`r colorize("Jaccard similarity decreased", "blue")` in unprotected areas only, with no trend for protected areas.
We found significant dietary shifts across communities, with a significant increasing general trend in fish consumption, which was also reflected in increasing fish consumption trends for bird studies and temperate studies.
`r colorize("Vertebrate consumption significantly declined", "blue")` in marine studies and tropical studies.
The trend for marine vertebrate consumption was significantly more negative than for terrestrial studies and the temperate trend was significantly more negative than tropical studies.

At the study level, 11 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
For observed functional diversity metrics, 11 of 50 studies exhibited a trend in at least one metric, and 7 of 50 studies exhibited a significant trend for at least one corrected metric (Table \ref{tab:trendTab}).
In general, there were more significant trends for observed functional diversity metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
Study-level slopes were not significantly related to start year or duration of time series for any metric.

@@

We assessed the sensitivity of general trend results `r colorize("for all metrics", "blue")` to major data processing decision by rerunning models with increasingly conservative subsets of the data.
Trends for Jaccard similarity and fish consumption were not sensitive to either time series duration or trait coverage.
After excluding time series with less than three years we found an increasing trend for body mass that remained after excluding time series of less than four and five years.
The body size trend did however disappear after excluding time series of less then 10 years.
A complete list of models run in the sensitivity analysis and their results can be found in the supplement.

@@

Category-specific trends in `r colorize("species", "blue")` turnover highlight the types of communities experiencing change in community composition.
For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall `r colorize("turnover", "blue")` trend while disguising relative stasis in marine bird and mammal communities.
We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.

Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact.
For example, we found no trend in mean body mass in bird or mammal communities.
In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size.
This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore reestablishment are overshadowing climate change impacts[@hantak2021], or current shifts are happening at an intraspecific level not captured by our data.
Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
